---
title: Exploring Data
bibliography: references.bib
link-citations: true
execute:
  output: false
---

```{r}
#| label: load-packages
library(tidyverse)
```

# Exploring BC Air Quality Data

The hourly PM2.5 observations wer obtained from the National Air Pollution Surveillance (NAPS) Program [@ecccNAPS2026].

## Inspect Air Quality Datasets
```{r}
#| label: load-data
pm25_2023 <- read_csv("data/raw/airquality/2023/data_2023.csv")
stations_2023 <- read_csv("data/raw/airquality/2023/stations_2023.csv")
n_distinct(pm25_2023$NAPS_ID) # monitoring stations
```

**data_2023.csv:**  
Data has 481800 hourly PM2.5 observations across BC for 2023 (8 columns). Each row shows one PM2.5 concentration measurement, one monitoring station, one date, start, and end time.  
* Important: NAPS_ID -> can connect to stations_2023.csv

**stations_2023.csv:**  
Observations are recorded by **55** monitoring stations.  
Useful fields out of 43:  

- NAPS_ID  
- Station_Name  
+ City  
+ Latitude: will be used to calculate distances from nearby fires  
+ Longitude: same as latitude  
+ Elevation_m  
+ Site_Type  
+ Urbanization  
+ Land_Use  
+ AQMS_Airzone  


### Research
Particulate Matter (PM) is a complex mix of solids and aerosols. PM2.5 describes the fine particulate matter that are 2.5 microns or less in diameter (PM2.5).

**Ambient Air Quality Standard in BC:**  
24-Hour Average: 25 µg/m^3  
Annual: 8 µg/m^3  [@bcPM25Objectives2026]

## Joined Station and PM2.5

```{r}
# check that the number and IDs of monitoring stations match
nrow(stations_2023)
n_distinct(stations_2023$NAPS_ID)

anti_join(
  pm25_2023 |> distinct(NAPS_ID),
  stations_2023 |> distinct(NAPS_ID),
  by = "NAPS_ID"
)
```

```{r}
#| label: clean-stations
station_info_2023 <- stations_2023 |>
  select(
    NAPS_ID,
    Station_Name,
    City,
    Latitude,
    Longitude,
    Elevation_m,
    Site_Type,
    Urbanization,
    Land_Use,
    AQMS_Airzone
  )
glimpse(station_info_2023)
```

```{r}
#| label: join
pm25_2023_joined <- pm25_2023 |>
  left_join(station_info_2023, by = "NAPS_ID")
glimpse(pm25_2023_joined) # result should have 10+8-1=17 columns
```

### 1. How much is PM2.5 data missing?
```{r}
pm25_2023_joined |>
  summarise(
    total_rows = n(),
    missing_concentration = sum(is.na(Concentration)),
    percent_missing = mean(is.na(Concentration)) * 100
  )
```
27,784 rows missing concentration (5.77%)

### 2. PM2.5 concentration distribution
```{r}
#| output: true
summary(pm25_2023_joined$Concentration)
```

Max of 2112 is very big compared to others.
```{r}
#| label: compare-2112
pm25_2023_joined |>
  filter(!is.na(Concentration)) |>
  arrange(desc(Concentration)) |>
  select(
    Date_yyyy_mm_dd,
    Start_Time_Local_Standard,
    Concentration,
    Station_Name,
    City,
    NAPS_ID
  ) |>
  slice_head(n = 20)
```
```{r}
#| label: summary-of-high
pm25_2023_joined |>
  summarise(
    zero = sum(Concentration == 0, na.rm=TRUE),
    negative = sum(Concentration < 0, na.rm=TRUE),
    above_100 = sum(Concentration > 100, na.rm=TRUE),
    above_500 = sum(Concentration > 500, na.rm=TRUE),
    above_1000 = sum(Concentration > 1000, na.rm=TRUE)
  )
```
Here, though 2112 is an extreme value, the surrounding observations show sustained high concentrations at the same station over consecutive hours. Therefore, it should not be removed as an erroneous outlier. Note: Fort St. John Max PM2.5 - 2112 and Kamloops Max PM2.5 - 982.

### 3. Are Pollutant and Units actually constant?
```{r}
pm25_2023_joined |>
  count(Pollutant)

pm25_2023_joined |>
  count(Units)
```
Units have 27,784 missing values and 454,016 rows as µg/m^3 while Pollutant is constant
Note: 27,784 rows of Units = 27,784 rows of Concentration are N/A
```{r}
pm25_2023_joined |>
  summarise(
    mismatched_missing = sum(is.na(Concentration) != is.na(Units))
  )
```
Result of 0 verifies that NAPS leaves Units blank when PM2.5 is missing.

### 4. Missing Data by Station
```{r}
pm25_2023_joined |>
  group_by(NAPS_ID, Station_Name) |>
  summarise(
    observations = n(),
    missing = sum(is.na(Concentration)),
    percent_missing = mean(is.na(Concentration)) * 100
  ) |>
  arrange(desc(percent_missing))
```
Quite high for few ie. Powell River... has 40.8% missing, and the total missing is 5.77%. A station like Powell River might distort station specific analyses.  

Using a heat map, check whether missing is a random hourly equipment gaps throughout year (less concerning) or an entire month worth of data is missing (concerning).

*Figure 1: Missing PM2.5 Measurements by Station and Month*  
Stations have missingness in subsequent months. Thus, it is not just random hourly equipment gaps. For example, Powell River Station has a high concentration of missing values from August to December.
```{r}
# separate month from date and calculate percent missing
pm25_missing_monthly <- pm25_2023_joined |>
  mutate(Month = lubridate::month(Date_yyyy_mm_dd, label = TRUE)) |>
  group_by(Station_Name, Month) |>
  summarise(
    percent_missing = mean(is.na(Concentration)) * 100,
    .groups = "drop"
  )
```
```{r}
#| label: to-save
#| echo: false
pm25_missing_monthly_plot <- pm25_missing_monthly |>
  ggplot(aes(x = Month, y = Station_Name, fill = percent_missing)) +
  geom_tile() +
  labs(
    title = "Missing PM2.5 Heat Map by Station and Month",
    x = "Month",
    y = "Monitoring station",
    fill = "Missing (%)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.7),
    axis.text.x = element_text(angle = 45, size = 8),
    axis.text.y = element_text(size = 5)
  )
```
```{r}
#| label: to-show
#| output: true

# showing only the top 
top_missing_stations <- pm25_2023_joined |>
  group_by(Station_Name) |>
  summarise(
    percent_missing = mean(is.na(Concentration)) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(percent_missing)) |>
  slice_head(n = 15) |>
  pull(Station_Name)
pm25_missing_monthly_top <- pm25_missing_monthly |> filter(Station_Name %in% top_missing_stations)

pm25_missing_monthly_top_plot <- pm25_missing_monthly_top |>
  ggplot(aes(
    x = Month,
    y = fct_reorder(Station_Name, percent_missing, .fun = max),
    fill = percent_missing
  )) +
  geom_tile() +
  labs(
    title = "Monthly Missing PM2.5 for Stations with the Highest Missingness",
    subtitle = "British Columbia, 2023",
    x = "Month",
    y = "Monitoring station",
    fill = "Missing (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 7),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5))
pm25_missing_monthly_top_plot
```

---

# Exploring BC Wildfire Data

The wildfire data were obtained from the Candian National Fire Database (CNFDB) [@cnfdb2026]. This dataset contains vegetation fire records and only provides data on "large fires" of at least 200 hectares.

## Inspect Data
```{r}
large_fires <- read_csv("data/raw/wildfire/NFDB_point_large_fires.txt")
glimpse(large_fires)
problems(large_fires) 

# fix wrong predetermined type
large_fires <- read_csv(
  "data/raw/wildfire/NFDB_point_large_fires.txt",
  col_types = cols(ATTK_DATE = col_datetime()))
problems(large_fires)

#| output
head(large_fires)
```

## 2022-2024 BC Fires
```{r}
# large fires in BC during 2022-2024
large_fires |> count(SRC_AGENCY, sort=TRUE)
large_fires_bc <- large_fires |>
  filter(SRC_AGENCY == "BC", YEAR %in% 2022:2024)

glimpse(large_fires_bc)
```

Total BC Large Fires (1950-2025): 3209  
Total BC Large Fires (2022-2024): 382

```{r}
#| output: true
large_fires_bc |>
  select(REP_DATE) |>
  arrange(REP_DATE) |>
  head()

large_fires_bc |>
  count(YEAR)

large_fires_bc |>
  count(FIRE_TYPE, sort=TRUE)

large_fires_bc |>
  count(CAUSE22, sort=TRUE)
```
REP_DATE arrangement shows that the first observation of wildfire in BC (2022-2024) starts on May 31st, 2022.

Year shows 2023 has the greatest amount of fires.

FIRE_TYPE does not provide useful information for BC data.

From the NFDB_point_metadata.pdf,
CAUSE:  
[N] Natural/Lightning caused  
[H] Human caused  
[U] Unknown cause  
* Remember to change these for processed data

*Figure 2: Large fires by year*  
From BC large fires by year 2022-2024, 2023 had a much more fires than in other two years. 2023 was also the year when the extreme values of PM2.5 appeared.  
```{r}
#| label: bar-graph
#| output: true
large_fires_bc |>
  count(YEAR) |>
  ggplot(aes(x = factor(YEAR), y = n)) +
  geom_col() +
  labs(
    title = "Number of Large Fire Records in BC by Year",
    x = "Year",
    y = "Number of large fires"
  ) +
  theme_minimal()
```

```{r}
#| label: missing-values
#| output: true
large_fires_bc |>
  summarise(
    fires = n(),
    missing_latitude = sum(is.na(LATITUDE)),
    missing_longitude = sum(is.na(LONGITUDE)),
    missing_date = sum(is.na(REP_DATE)),
    missing_size = sum(is.na(SIZE_HA))
  )
```

Missing values of 0 is great as latitude, longitude, rep_date, size_ha will be useful in calculating distances between fire and the station.  

```{r}
#| output: true
large_fires_bc |>
  count(PRESCRIBED, sort=TRUE)

large_fires_bc |>
  count(MORE_INFO, sort=TRUE)

large_fires_bc |>
  count(RESPONSE, sort = TRUE)
```

We can see that PRESCRIBED gives non-usable information for BC records. MORE_INFO shows description of location having 281 different values. RESPONSE (from NFDB_points_metadata.pdf) has three classes: [FUL] Full response, [MOD] Modified response, [MON] Monitor.

```{r}
#| label: area
#| output: true
large_fires_bc |>
  summarise(
    min_size_ha = min(SIZE_HA),
    median_size_ha = median(SIZE_HA),
    mean_size_ha = mean(SIZE_HA),
    max_size_ha = max(SIZE_HA)
  )
```

```{r}
#| output: true
#| echo: false
fire_size_by_year <- large_fires_bc |>
  group_by(YEAR) |>
  summarise(
    fires = n(),
    min_size_ha = min(SIZE_HA),
    median_size_ha = median(SIZE_HA),
    mean_size_ha = mean(SIZE_HA),
    max_size_ha = max(SIZE_HA)
  )

# create a table
knitr::kable(
  fire_size_by_year,
  col.names = c("Year", "Number of Fires", "Minimum Size(ha)", "Median Size(ha)", "Mean Size(ha)", "Maximum Size(ha)"),
  digits = 1,
  caption = "Figure 3: Summary of BC Large Fire Sizes by Year"
)
```


*Figure 4: Wildfire Sizes Box Plot by Year*  
Table and a box plot. With median of 1027 and mean size of 10550, we can understand that it is strongly right-skewed. 

```{r}
#| output: true
large_fires_bc |>
  ggplot(aes(
    x = factor(YEAR),
    y = SIZE_HA
  )) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Figure 4: Distribution of Large Fire Size in BC",
    x = "Year",
    y = "Fire size (hectares, log scale)"
  ) +
  theme_minimal()
```
The logarithmic scaling is useful when there are both very large and very small values, squishing the large ones and spacing the small ones.

**Summary of Data:** 382 BC large vegetation fire records ($\geq$ 200ha) from the Canadian National Fire Database for 2022-2024

---

# Explore 2022 & 2024 Air Quality Data
## Inspect
Repeating what was done for 2023 is tedious and unnecessary.

First, load.
```{r}
pm25_2022 <- read_csv("data/raw/airquality/2022/data_2022.csv")

stations_2022 <- read_csv("data/raw/airquality/2022/stations_2022.csv")

pm25_2024 <- read_csv("data/raw/airquality/2024/data_2024.csv")

stations_2024 <- read_csv("data/raw/airquality/2024/stations_2024.csv")
```

Second, check if the column names are same.
```{r}
#| output: true
identical(names(pm25_2022), names(pm25_2023))
identical(names(pm25_2023), names(pm25_2024))

identical(names(stations_2022), names(stations_2023))
identical(names(stations_2023), names(stations_2024))
```

Third, all three years together.
```{r}
#| label: overview
#| output: true
tibble(
  Year = c(2022, 2023, 2024),

  PM25_Rows = c(
    nrow(pm25_2022),
    nrow(pm25_2023),
    nrow(pm25_2024)
  ),
  
  Stations_With_PM25 = c(
    n_distinct(pm25_2022$NAPS_ID),
    n_distinct(pm25_2023$NAPS_ID),
    n_distinct(pm25_2024$NAPS_ID)
  ),
  
  Stations_In_Metadata = c(
    nrow(stations_2022),
    nrow(stations_2023),
    nrow(stations_2024)
  )
)

pm25_missing_years <- bind_rows(
    pm25_2022 |> mutate(Year = 2022),
    pm25_2023 |> mutate(Year = 2023),
    pm25_2024 |> mutate(Year = 2024)
  ) |>
  group_by(Year) |>
  summarise(
    observations = n(),
    missing = sum(is.na(Concentration)),
    percent_missing = mean(is.na(Concentration)) * 100
  )
pm25_missing_years
```
Stations: 56 -> 55 -> 58 (changing number)
Missing: 4.84 -> 5.77 -> 7.23 (growing)

Note: 2024 is a leap year so it has much more rows

Final, combine.
```{r}
pm25_all <- bind_rows(
  pm25_2022,
  pm25_2023,
  pm25_2024
)
```

---

# How to determine whether a fire was relevant to PM2.5?

REP_DATE: Date associated with fire  
ATTK_DATE: Fire suppression start date  
OUT_DATE: Fire out date  

REP_DATE == PM2.5 date would not work because fire can still be burning on other days and make the air quality bad. Plus, wind blows and the PM2.5 cover areas in varying times.

```{r}
#| label: missing
#| output: true
large_fires_bc |>
  summarise(
    total_fires = n(),
    missing_rep_date = sum(is.na(REP_DATE)),
    missing_attack_date = sum(is.na(ATTK_DATE)),
    missing_out_date = sum(is.na(OUT_DATE))
  )
```

ATTK_DATE data is unusable since all of them are missing.

```{r}
#| label: get-a-sample
large_fires_bc |>
  select(
    FIRE_ID,
    FIRENAME,
    REP_DATE,
    ATTK_DATE,
    OUT_DATE,
    SIZE_HA,
    RESPONSE,
    MORE_INFO
  ) |>
  slice_head(n = 20)
```

## Using OUT_DATE

Calculating the duration of the fire in days by finding the difference between reported dates of fire and the date the fire was put out.

```{r}
#| output: true

# calculate duration through OUT_DATE
large_fires_bc |>
  filter(!is.na(OUT_DATE)) |>
  mutate(duration_days = as.numeric(difftime(OUT_DATE, REP_DATE, units = "days"))) |>
  summarise(
    min_duration = min(duration_days),
    median_duration = median(duration_days),
    mean_duration = mean(duration_days),
    max_duration = max(duration_days)
  )
```

The min_duration of -2 shows that there are inconsistencies.

```{r}
#| output: true

# find out which row gives a duration days below 0
large_fires_bc |>
  filter(!is.na(OUT_DATE)) |>
  mutate(duration_days = as.numeric(difftime(OUT_DATE, REP_DATE, units = "days"))) |>
  filter(duration_days < 0) |>
  select(
    NFDBFIREID,
    FIRENAME,
    REP_DATE,
    OUT_DATE,
    duration_days,
    SIZE_HA,
    MORE_INFO
  )
```

Only one row. This can be treated as missing when cleaning data.

Finding the PM2.5 data between REP_DATE and OUT_DATE is very crude and potentially treat a fire equally as day 2 as day 92.

## Using REP_DATE only

REP_DATE has no missing values making it more reliable. 

This uses a recent-fire variable (to be created). It is going to check previous 7/14/30 days of air-quality stations and find any large fires within different perimeters in those days.

This is not looking for currently "active" fires like was told before.

---

# Conclusion

## Air Quality Data

- NAPS provides complete hourly station grids for each year
- PM2.5 measurement missingness ranges from 4.84% to 7.23% across 2022–2024
- Missingness varies substantially between monitoring stations and many appear in subsequent months
- Extreme PM2.5 values occur in sustained episodes

## Large Fire Data

- 382 BC large vegetation fire records available from 2022-2024
- Fire size is strongly right-skewed
- Following provides non-usable information for BC 2022-2024 records: PRESECRIBED, ATTK_DATE, FIRE_TYPE
- OUT_DATE contains one inconsistent record that was found
- Use recent-fire windows based on `Fire_Date` to connect to PM2.5 date

## Analysis Design
One monitoring station on one day with columns being:

- Date
- NAPS_ID
- Station_Name
- Latitude
- Longitude
- mean_pm25
- max_pm25
- hours_observed
- recent_fire_count_100km
- recent_fire_count_250km
- recent_fire_area_250km
- nearest_recent_fire_km
- Year
- Month

These are the questions I have formed and will be answering:  

- How did daily PM2.5 differ between 2022-2024?
- Can we classify a station day as High-PM2.5 vs. Normal?
- Can recent large-fire proximity/activity help predict daily PM2.5?
- Do distinct types of station-days apppear naturally?