---
title: "Build Analysis Dataset"
bibliography: references.bib
link-citations: true
execute:
    output: false
---

Transforms the raw datasets to processed ones.

```{r}
#| label: load
library(tidyverse)
library(geosphere)
```

The table below summarizes the most important variables used throughout the data-building and analysis workflow.  

```{r}
#| label: variable-reference
#| echo: false
#| output: true
variable_reference <- tibble(
  `Air Quality` = c(
    "Station_Date",
    "NAPS_ID",
    "Station_Name",
    "Latitude",
    "Longitude",
    "mean_pm25",
    "max_pm25",
    "hours_observed",
    "valid_day"
  ),

  Fire = c(
    "Fire_ID",
    "Fire_Year",
    "Report_Date",
    "Fire_Date",
    "Fire_Latitude",
    "Fire_Longitude",
    "Size_Ha",
    "Cause",
    "Response"
  ),

  Relationship = c(
    "days_since_fire",
    "Distance_km",
    "fire_count_100km_{7/14/30}d",
    "fire_count_250km_{7/14/30}d",
    "fire_area_250km_{7/14/30}d",
    "nearest_fire_km_{7/14/30}d",
    "",
    "",
    ""
  )
)

knitr::kable(
  variable_reference,
  caption = "Key variables used throughout the analysis",
  na = ""
)
```

# Prepare Air Quality Data

First, load data.

```{r}
#| label: load-air-quality-data
pm25_2022 <- read_csv(
  "data/raw/airquality/2022/data_2022.csv",
  show_col_types = FALSE)

stations_2022 <- read_csv(
  "data/raw/airquality/2022/stations_2022.csv",
  show_col_types = FALSE)

pm25_2023 <- read_csv(
  "data/raw/airquality/2023/data_2023.csv",
  show_col_types = FALSE)

stations_2023 <- read_csv(
  "data/raw/airquality/2023/stations_2023.csv",
  show_col_types = FALSE)

pm25_2024 <- read_csv(
  "data/raw/airquality/2024/data_2024.csv",
  show_col_types = FALSE)

stations_2024 <- read_csv(
  "data/raw/airquality/2024/stations_2024.csv",
  show_col_types = FALSE)
```

Second, select station data columns.

```{r}
#| label: select-station-columns

station_columns <- c(
  "NAPS_ID",
  "Station_Name",
  "City",
  "Latitude",
  "Longitude",
  "Elevation_m",
  "Site_Type",
  "Urbanization",
  "Land_Use",
  "AQMS_Airzone"
)

station_info_2022 <- stations_2022 |>
  select(all_of(station_columns))

station_info_2023 <- stations_2023 |>
  select(all_of(station_columns))

station_info_2024 <- stations_2024 |>
  select(all_of(station_columns))
```

Third, join station and pm25 data by year.

```{r}
#| label: join-station-data

# join each year and add Year column
pm25_2022_joined <- pm25_2022 |>
    left_join(station_info_2022, by = "NAPS_ID") |>
    mutate(Year = 2022)

pm25_2023_joined <- pm25_2023 |>
    left_join(station_info_2023, by = "NAPS_ID") |>
    mutate(Year = 2023)

pm25_2024_joined <- pm25_2024 |>
    left_join(station_info_2024, by = "NAPS_ID") |>
    mutate(Year = 2024)
pm25_2022_joined
```

Last, combine all years.

```{r}
#| label: combine-years

pm25_joined <- bind_rows(
    pm25_2022_joined,
    pm25_2023_joined,
    pm25_2024_joined
)
glimpse(pm25_joined)
pm25_joined |> count(Year)
```

Verification.
```{r}
#| label: verify-join-row-counts
#| output: true

tibble(
    Year = c(2022, 2023, 2024),
    Before_Join = c(
        nrow(pm25_2022),
        nrow(pm25_2023),
        nrow(pm25_2024)
    ),
    After_Join = c(
        nrow(pm25_2022_joined),
        nrow(pm25_2023_joined),
        nrow(pm25_2024_joined)
    )
)
```

Before_Join and After_Join values are equal, verifying that the joining worked as intended.

### Valid Days

Environment and Climate Change Canada considers a PM2.5 daily 24-hour average valid when at least 75% of the hourly measurements (18 to 24) are available [@ecccAirQuality2026]. Therefore, station-days with fewer than 18 observed hours are retained for data-quality tracking but their daily PM2.5 sumaries are treated as missing.

- hours_observed: number of non-missing hourly readings
- mean_pm25: daily 24-hour average
- max_pm25: largest observed hourly concentration of the day
- Station_Date: change name of Date_yyyy_mm_dd
- Month: separate month
- valid_day: TRUE or FALSE

```{r}
#| label: station-day
pm25_daily <- pm25_joined |>
  group_by(
    Year,
    Date_yyyy_mm_dd,
    NAPS_ID,
    Station_Name,
    City,
    Latitude,
    Longitude,
    Elevation_m,
    Site_Type,
    Urbanization,
    Land_Use,
    AQMS_Airzone
  ) |>
  summarise(
    # collapse hourly observations
    hours_observed = sum(!is.na(Concentration)), 

    # collapse hourly observation of pm25 into a mean, max pm25 of the day
    mean_pm25 = if (sum(!is.na(Concentration)) >= 18) {
        mean(Concentration, na.rm=TRUE)
    } else {
        NA_real_
    },
    max_pm25 = if (sum(!is.na(Concentration)) >= 18) {
        max(Concentration, na.rm=TRUE)
    } else {
        NA_real_
    },
    .groups = "drop"
  )

pm25_daily <- pm25_daily |>
    rename(
        Station_Date = Date_yyyy_mm_dd
    ) |>
    mutate(Month = lubridate::month(Station_Date), valid_day = hours_observed >= 18)

glimpse(pm25_daily)
pm25_daily |> count(Year, valid_day)
```

Do not filter valid_day = FALSE as this is still useful information.

```{r}
#| output: true
# percent
pm25_daily |>
  group_by(Year) |>
  summarise(
    station_days = n(),
    valid_days = sum(valid_day),
    invalid_days = sum(!valid_day),
    percent_valid = mean(valid_day) * 100,
    .groups = "drop"
  )
```

### Hours Observed
```{r}
#| label: hour-observed-distribution
#| output: true
pm25_daily_hours_observed <- pm25_daily |>
    count(hours_observed) |>
    arrange(desc(hours_observed))

pm25_daily |> 
    ggplot(aes(x = hours_observed)) +
    geom_bar() +
    scale_x_continuous(breaks = 0:24) +
    labs(
      title = "Figure 1: Distribution of Hours Observed",
      x = "Hours Observed",
      y = "Count"
    ) +
    geom_vline(xintercept = 17.5, linetype = "dashed") +
    theme_minimal()
```

---

# Prepare Wildfire Data

- Filter 2022-2024 BC
- Don't use: ATTK_DATE, PRESCRIBED, FIRE_TYPE (+ unused: PROTZONE, etc.)
- Inconsistent: OUT_DATE
- Rename LATITUDE and LONGITUDE since it is also in stations
- Recode N,H,U in CAUSE22 and FUL,MOD,MON in RESPONSE
- REP_DATE show only date and not time for usage

First, load.
```{r}
#| label: load-large-fires-data

large_fires <- read_csv(
  "data/raw/wildfire/NFDB_point_large_fires.txt",
  col_types = cols(ATTK_DATE = col_datetime()))
glimpse(large_fires)
```

Second, clean.
```{r}
#| label: clean-large-fires-data

large_fires_clean <- large_fires |>
  filter(
    SRC_AGENCY == "BC",
    YEAR %in% 2022:2024
  ) |>
  mutate(
    Fire_Year = YEAR,
    Report_Date = as.Date(REP_DATE),
    Report_Year = lubridate::year(Report_Date),
    Cause = recode(
      CAUSE22,
      "N" = "Natural",
      "H" = "Human",
      "U" = "Unknown"
    ),
    Response = recode(
      RESPONSE,
      "FUL" = "Full",
      "MOD" = "Modified",
      "MON" = "Monitor"
    )
  ) |>
  select(
    Fire_ID = NFDBFIREID,
    Fire_Year,
    Report_Date,
    Report_Year,
    Fire_Latitude = LATITUDE,
    Fire_Longitude = LONGITUDE,
    Size_Ha = SIZE_HA,
    Cause,
    Response
  )
glimpse(large_fires_clean)
```
Now, it went from 24 variables to 9 variables.

Third, verify.
```{r}
#| output: true
summary(large_fires_clean$Size_Ha)

large_fires_clean |>
  count(Fire_Year, Report_Year)
```

Size_Ha result matches the one from exploration. From the result, we find that there are 4 rows disagreeing on the year.

### Resolving conflicting data

```{r}
#| output: true
large_fires_clean |>
  filter(Fire_Year != Report_Year) |>
  select(
    Fire_ID,
    Fire_Year,
    Report_Date,
    Report_Year,
    Size_Ha,
    Cause,
    Response
  )
```

- BC-2024-2024-G90228: Nogah Creek - May 5, 2024 [@bcwsG90228]
- BC-202402024-G90399: CAN'T FIND
- BC-2024-2024-G90777: (Unverified source) Burn severity report, 2024
- BC-2024-2024-G90207: Party Creek - May 2, 2024 [@bcwsG90207]

*Important: Not to overwrite Report_Date; instead, create a new variable: Fire_Date


```{r}
# two fire dates identified, two fire given NA
large_fires_clean <- large_fires_clean |>
  mutate(
    Fire_Date = case_when(
      Fire_ID == "BC-2024-2024-G90228" ~ as.Date("2024-05-05"),
      Fire_ID == "BC-2024-2024-G90207" ~ as.Date("2024-05-02"),

      Fire_ID %in% c("BC-2024-2024-G90399", "BC-2024-2024-G90777") ~ as.Date(NA),

      TRUE ~ Report_Date
    )
  )

# add Date_Status to indicate the corrected and unresolved ones
large_fires_clean <- large_fires_clean |>
  mutate(
    Date_Status = case_when(
      Fire_ID %in% c("BC-2024-2024-G90228", "BC-2024-2024-G90207") ~ "Verified correction",
      Fire_ID %in% c("BC-2024-2024-G90399", "BC-2024-2024-G90777") ~ "Unresolved",
      TRUE ~ "Original"
    )
  )

glimpse(large_fires_clean)
```

With this addition of variables, the total became 11.

Previously, since it had no missing values, REP_DATE was assumed to be a reliable field. Yet, this showed that everything needs to be heeded to carefully, including validity and consistency.

```{r}
#| label: last-verification
#| output: true
large_fires_clean |>
  count(Date_Status)

large_fires_clean |>
  summarise(
    usable_temporal_records = sum(!is.na(Fire_Date)),
    unusable_temporal_records = sum(is.na(Fire_Date))
  )
```

---

# Spatial-Distance Calculations

Using both the PM2.5 joined data (pm25_daily) and large wildfires data (large_fires_clean)

## 1. Checking
Check that the station's coordinates does not change.
```{r}
#| label: check-station-coordinates
#| output: true

station_coordinate_check <- pm25_daily |>
  distinct(NAPS_ID, Latitude, Longitude) |>
  count(NAPS_ID, name = "coordinate_versions") |>
  filter(coordinate_versions > 1)

station_coordinate_check
```

The result confirms that the each station's coordinates did not change.

For this, going to use geosphere.   
*Important: geosphere expects coordinates in c(Longitude, Latitude) order

## 2. Take station locations from Air Quality data

Extract station ID, name, city, and location coordinates.
```{r}
#| label: station-locations
#| output: true

station_locations <- pm25_daily |>
  distinct(
    NAPS_ID,
    Station_Name,
    City,
    Latitude,
    Longitude
  )

glimpse(station_locations)
nrow(station_locations)
```

Across the three years, there were stations that only appear in only one or two years.

## 3. Make every station/fire pair
```{r}
#| label: usable-temporal-fire-data
# create a dataframe of only usable data
large_fires_temporal <- large_fires_clean |>
  filter(!is.na(Fire_Date))

glimpse(large_fires_temporal)
```

AI was used to suggest using `tidyr::crossing()` to do cartesian product.
```{r}
#| label: station-fire-pairs
# use cartesian product
station_fire_pairs <- tidyr::crossing(
  station_locations,
  large_fires_temporal |>
    select(
      Fire_ID,
      Fire_Date,
      Fire_Latitude,
      Fire_Longitude,
      Size_Ha
    )
)
glimpse(station_fire_pairs)
```
Results in 10 variables and 22,800 data.

### AI Audit

What I want:  
Station A | data  
Station A | data  
Station A | data  
...(until it runs through large_fires_temporal rows)   
Station B | data  
...(same thing)  

This means that each station is associated with all the data in large_fires_temporal. Thus, the following MUST BE TRUE:
```{r}
#| output: true
nrow(station_fire_pairs) == nrow(station_locations) * nrow(large_fires_temporal)
```

### 4. Calculate distance

**AI disclosure:** ChatGPT was used to suggest using `geosphere::distHarversine` for geographic distance calculation. The function's behaviour and assumptions were verified using geosphere documentations [@geosphere2026].
```{r}
#| label: station-fire-distances

station_fire_distances <- station_fire_pairs |>
  mutate(
    Distance_km =
      geosphere::distHaversine(
        cbind(Longitude, Latitude),
        cbind(Fire_Longitude, Fire_Latitude)
      ) / 1000
  )
```

### AI-Audit

Latitude and longitude represent angular geographic coordinates rather than distances in kilometres. Therefore, directly calculating Euclidean distance from differences in latitude and longitude cannot work.  

distHaversine() function calculates the shortest distance between two points. It also accepts longitude and latitude coordinates and calculate spherically [@geosphere2026]. Ex. distHarversine(c(0,0), c(90,90)). It returns the distance in metres which is the reason for /1000 to convert to kilometres.

Try a known value: following must be 111.3 km.

```{r}
#| output: true
geosphere::distHaversine(
  c(0, 0),
  c(1, 0)
) / 1000
```
```{r}
#| output: true
# check distances
summary(station_fire_distances$Distance_km)

station_fire_distances |>
  arrange(Distance_km) |>
  select(
    NAPS_ID,
    Station_Name,
    Fire_ID,
    Fire_Date,
    Size_Ha,
    Distance_km
  ) |>
  slice_head(n = 10)
```

---

# Collapsing Fire Data

## 1. Time filtering

Checking if a fire have been active within 30 days of a PM2.5 data. If so, then register as a candidate.
```{r}
# calculates 30 days before Station_Date
station_day_fire_candidates <- pm25_daily |>
  mutate(Window_Start = Station_Date - lubridate::days(30))
# check
station_day_fire_candidates |>
  slice_sample(n=5) |> 
  select(Window_Start, Station_Date)
```

Join station_fire_distances and station_day_fire_candidates. Comparing variables of the two datasets:
```{r}
#| label: variable-comparison
all_columns <- union(names(station_fire_distances), names(station_day_fire_candidates))

data.frame(
  station_fire_distances = ifelse(
    all_columns %in% names(station_fire_distances),
    all_columns,
    NA
  ),
  station_day_fire_candidates = ifelse(
    all_columns %in% names(station_day_fire_candidates),
    all_columns,
    NA
  ),
  existInBoth = all_columns %in% names(station_fire_distances) &
                all_columns %in% names(station_day_fire_candidates)
)
```

We are selecting `NAPS_ID`, `Fire_ID`, `Fire_Date`, `Size_Ha`, `Distance_km` from the station-fire distances dataframe. Then joining it when each of its data matches with each of the data in fire-candidates-to-station dataframe.

The match is determined by the following conditions: 

- NAPS_ID
- if Fire_Date (in station-fire-dist) is over or equal to Window_Start (in fire-candidate) 
- if Fire_Date (in station-fire-dist) is before or equal to Station_Date (in fire-candidate)

```{r}
station_day_fire_candidates <- station_day_fire_candidates |>
  inner_join(
    station_fire_distances |> 
      select(NAPS_ID, Fire_ID, Fire_Date, Size_Ha, Distance_km),

    by = join_by(
      NAPS_ID, 
      Window_Start <= Fire_Date, 
      Station_Date >= Fire_Date
    ),
    relationship = "many-to-many" # since each station-day can have many fire
  ) |>
  mutate(days_since_fire = as.numeric(Station_Date - Fire_Date))

glimpse(station_day_fire_candidates)
```

```{r}
#| label: audit-temporal-join
#| output: true

summary(station_day_fire_candidates$days_since_fire)

all(
  station_day_fire_candidates$days_since_fire >= 0 &
  station_day_fire_candidates$days_since_fire <= 30
)
```

These verify that the `days_since_fire` falls between 0 to 30.

### What does station_day_fire_candidates actually show?
A data: one station-day paired with one fire reported in the previous 30 days

For example, on the same date on the same station, there will be many fire associated with it.
```{r}
#| output: true
station_day_fire_candidates |>
  arrange(Station_Date, Station_Name) |>
  select(Station_Date, Station_Name, Fire_ID, days_since_fire, Distance_km) |>
  head()
```

Example of an extreme PM2.5 events (Kamloops case - found in file 01). It shows that there really was a cluster of large-fire reports during one period.
```{r}
#| echo: false
station_day_fire_candidates |>
  filter(
    Station_Date == as.Date("2023-08-21"),
    str_detect(Station_Name, "KAMLOOPS")
  ) |>
  arrange(Distance_km) |>
  select(
    Station_Date,
    Station_Name,
    Fire_ID,
    Fire_Date,
    days_since_fire,
    Size_Ha,
    Distance_km
  ) |>
  slice_head(n = 10)
```

```{r}
#| label: kamloops-data-focus-aug-21
#| output: true
station_day_fire_candidates |>
  filter(
    Station_Date == as.Date("2023-08-21"),
    str_detect(Station_Name, "KAMLOOPS")
  ) |>
  arrange(Distance_km) |>
  transmute(
    Fire_ID,
    Fire_Date,
    days_since_fire,
    Size_Ha,
    Distance_km = round(Distance_km, 1)
  ) |>
  slice_head(n = 10)
```

## 2. Aggregating fire features

Currently, Kamloops station on August 21 might appear 20+ times because it can be paired with 20+ different fires. Instead, we want one data: Kamloops + 2023-0-21 to contain a SUMMARY of all relevant fires.

### Classify
First, create variables that classifies distances. When we ask: How many fires were within 100 km and 14 days?, we can answer simply.

```{r}
#| label: fire-window-indicators
station_day_fire_candidates <- station_day_fire_candidates |>
  mutate(
    within_7_days = days_since_fire <= 7,
    within_14_days = days_since_fire <= 14,
    within_30_days = days_since_fire <= 30,

    within_100_km = Distance_km <= 100,
    within_250_km = Distance_km <= 250
  )
```
Check if it works, must get TRUE for all.
```{r}
#| label: verify-fire-window-indicators
#| output: true
station_day_fire_candidates |>
  summarise(
    all_7_within_30 = all(!within_7_days | within_30_days),
    all_14_within_30 = all(!within_14_days | within_30_days),
    all_100_within_250 = all(!within_100_km | within_250_km)
  )

station_day_fire_candidates |>
  summarise(
    within_7_days = sum(within_7_days),
    within_14_days = sum(within_14_days),
    within_30_days = sum(within_30_days),

    within_100_km = sum(within_100_km),
    within_250_km = sum(within_250_km)
  )
```

```{r}
#| label: within-30-days-equal-to-row-count
#| echo: false
within_30_days_num <- station_day_fire_candidates |>
  filter(within_30_days) |>
  nrow()
nrow(station_day_fire_candidates) == within_30_days_num
```

### Aggregate
Second and importantly, collapse all these fire features into one row.
```{r}
#| label: aggregate-fire-features

station_day_fire_features <- station_day_fire_candidates |>
  group_by(NAPS_ID, Station_Date) |>
  summarise(
    # fire counts within 100 km
    fire_count_100km_7d = sum(within_100_km & within_7_days),
    fire_count_100km_14d = sum(within_100_km & within_14_days),
    fire_count_100km_30d = sum(within_100_km & within_30_days),

    # fire counts within 250 km
    fire_count_250km_7d = sum(within_250_km & within_7_days),
    fire_count_250km_14d = sum(within_250_km & within_14_days),
    fire_count_250km_30d = sum(within_250_km & within_30_days),

    # total size of nearby fires - for rough estimate of recently reported large-fire area nearby
    fire_area_250km_7d = sum(Size_Ha[within_250_km & within_7_days]),
    fire_area_250km_14d = sum(Size_Ha[within_250_km & within_14_days]),
    fire_area_250km_30d = sum(Size_Ha[within_250_km & within_30_days]),

    # nearest fire within each time window
    nearest_fire_km_7d =
      if (any(within_7_days)) {
        min(Distance_km[within_7_days])
      } else {
        NA_real_
      },
    nearest_fire_km_14d =
      if (any(within_14_days)) {
        min(Distance_km[within_14_days])
      } else {
        NA_real_
      },
    nearest_fire_km_30d = min(Distance_km),

    .groups = "drop"
  )

glimpse(station_day_fire_features)
```
```{r}
#| label: run-station-day-fire-features
#| output: true
station_day_fire_features |>
  arrange(desc(fire_count_250km_14d)) |>
  slice_head(n = 10)
```

Last, verify. The following must show 0. 
```{r}
#| label: run-no-duplicate-station-date
#| output: true
station_day_fire_features |>
  count(NAPS_ID, Station_Date) |>
  filter(n > 1)
```

This confirms that each row is has a unique station ID and station observed date as we intended: NAPS_ID + STATION_DATE + Summary of fires.

*Data Audit*  
fire_area_ is not reliable for prediction. The area of fire is always changing but this value is not since CNFDB dataset gives one size for each fire.

Auditing, following must all be TRUE.
```{r}
#| label: audit-fire-features
#| output: true

station_day_fire_features |>
  summarise(
    # count audit
    count_100_dist_250 = all(fire_count_100km_14d <= fire_count_250km_14d),
    count_7_days_14 = all(fire_count_250km_7d <= fire_count_250km_14d),
    count_14_days_30 = all(fire_count_250km_14d <= fire_count_250km_30d),

    # size audit
    area_7_days_14 = all(fire_area_250km_7d <= fire_area_250km_14d),
    area_14_days_30 = all(fire_area_250km_14d <= fire_area_250km_30d),

    # nearest value must be either same or less with greater data including the same ones
    nearest_30_days_14 = all(is.na(nearest_fire_km_14d) | nearest_fire_km_30d <= nearest_fire_km_14d)
  )
```

---

# Air Quality with Fire Features

Remember we joined: air quality station_locations and fire data. Now, we want every air-quality station day data preserved.

## Join the fire features to PM2.5
```{r}
#| label: join-fire-features

analysis_daily <- pm25_daily |>
  left_join(
    station_day_fire_features,
    by = c(
      "NAPS_ID",
      "Station_Date"
    )
  )

glimpse(analysis_daily)
```

If no qualifying fires exist, fire_count_ and fire_area_ shouldn't be NA; it should be 0. Yet, nearest_fire_distance_ are still NA because if it is 0, it would mean the fire is located at the monitoring station.

*AI Disclosure:* ChatGPT was used to find the code like `starts_with()` and `~ replace_na()`. It helped the code to be efficient as columns started with the same part and had to replace the same value: NA to 0.
```{r}
#| label: replace-na-0
analysis_daily <- analysis_daily |>
  mutate(
    across(
      starts_with("fire_count_"),
      ~ replace_na(.x, 0L)
    ),

    across(
      starts_with("fire_area_"),
      ~ replace_na(.x, 0)
    )
  )
```

Auditing,
```{r}
#| label: audit-replace-na-0
#| output: true

# check that it didn't lose rows
nrow(analysis_daily) == nrow(pm25_daily)

# check that the number NA values is 0
analysis_daily |>
  summarise(
    fire_count_ = sum(is.na(pick(starts_with("fire_count_")))),
    fire_area_ = sum(is.na(pick(starts_with("fire_area_"))))
  )
  
glimpse(analysis_daily)
```

Verified that no rows were lost (TRUE) and all NA values in fire_count_ and fire_area_ were replaced by 0.

---

# Real-World Cases

We will be looking through two stations that had a very high PM2.5 in 2023.

## Inspect the Kamloops & Fort St. John

We already have data about them from our file 01. Compare to check that every data have been preserved.

```{r}
#| label: case-studies
#| output: true

# Fort St. John
case_study_fires <- station_day_fire_candidates |>
  filter(
    (
      Station_Date == as.Date("2023-08-21") &
      str_detect(Station_Name, "KAMLOOPS")
    ) |
    (
      Station_Date == as.Date("2023-05-19") &
      str_detect(Station_Name, "FORT ST. JOHN")
    )
  ) |>
  filter(
    within_250_km,
    within_14_days
  ) |>
  select(
    Station_Date,
    Station_Name,
    Fire_ID,
    Fire_Date,
    days_since_fire,
    Size_Ha,
    Distance_km
  )
case_study_fires
```

### Audit: PM2.5 Daily Maximum
**Error**: For Kamloops and Fort St. John stations, `max_pm25` are supposed to be 982 and 2112 as we found in file 01, but this data shows that they are 596 and 461 -- equal to the `mean_pm25`.

Recalculating `max_pm25` and `mean_pm25` on raw-joined data (`pm25_joined`) showed that the raw data was preserved. Reviewing the aggregation code showed that `mean()` had accidentally been used for both variables -- a copy-past error. The `max_pm25` was corrected to use `max()`.

**Important**: Always compare processed results againt known raw observations. Code may execute successfully but will not gurantee that correctness.


## Validation

### Kamloops Station
For the validation, we will manually find out fire_count_250km_14d and compare with the actual value.

```{r}
#| label: manually-audit-kamloops-features
#| output: true

kamloops_manual_check <- station_day_fire_candidates |>
  filter(
    Station_Date == as.Date("2023-08-21"),
    str_detect(Station_Name, "KAMLOOPS")
  ) |>
  summarise(
    manual_count_250km_14d = sum(within_250_km & within_14_days),
    manual_area_250km_14d = sum(Size_Ha[within_250_km & within_14_days]),
    manual_nearest_fire_14d = min(Distance_km[within_14_days])
  )

kamloops_actual_values <- analysis_daily |>
  filter(
    Station_Date == as.Date("2023-08-21"),
    str_detect(Station_Name, "KAMLOOPS")
  ) |>
  select(
    fire_count_250km_14d,
    fire_area_250km_14d,
    nearest_fire_km_14d
  )

# manual
kamloops_manual_check

# actual
kamloops_actual_values
```

They are the same; validated. Plus, we can see that the area value is quite high.

---

# Final Processed Data

```{r}
#| label: save-analysis-data

write_csv(analysis_daily, "data/processed/analysis_daily.csv")

analysis_valid <- analysis_daily |>
  filter(valid_day)
write_csv(analysis_valid, "data/processed/analysis_valid.csv")

write_csv(large_fires_clean, "data/processed/large_fires_clean.csv")

write_csv(case_study_fires, "data/processed/case_study_fires.csv")
```

