---
title: Regressions
execute:
  output: false
---

```{r}
#| label: load-packages

library(tidyverse)
library(tidymodels)
library(patchwork)
```

```{r}
#| lable: load-data-and-seed
set.seed(555555)

analysis_valid <- read_csv(
  "data/processed/analysis_valid.csv",
  show_col_types = FALSE
)
```

---

# Question 1 - KNN Regression

*Among station days with a recently reported large fire, can fire proximity and recent fire activity be used to predict daily mean PM2.5?*


## Predictors

- nearest_fire_km_14d: how close was the nearest recent fire?
- fire_count_100km_7d: how much very nearby and very recent fires were there?
- fire_count_250km_14d: how much broader nearby fire activity was there?

This covers proximity, very local + very recent activities, and broader fire activities.

In other words it is essentially asking: *Which historical station days had a similar combination of nearest fire distance, nearby recent fire count, and broader recent fire count, and what PM2.5 did those days experience?*

```{r}
#| label: reg-pred
#| output: true
regression_data <- analysis_valid |>
  select(
    mean_pm25,
    nearest_fire_km_14d,  
    fire_count_100km_7d,
    fire_count_250km_14d
  )

# Check for NA values
regression_data |> summarise(across(everything(), ~ sum(is.na(.x))))

# filter NA values in nearest_fire_km_14d
nrow(regression_data)

regression_data <- analysis_valid |>
  filter(!is.na(nearest_fire_km_14d))

nrow(regression_data)
```

### Will regression work better
Classification threw away a lot of information making data analysis limited. However, regression retains the numerical data as is. Although this does not guarantee a good model, it seems to be better-fit for regression.

### Problem that might rise
Our target is extremely right-skewed and that may lead to a very large RMSE.

Plus, the predictors are still zero-heavy and the nearest fire in km within recent 14 days have many missing values. Due to lack of data, the regression might become unreliable.

---

## KNN Regression

### Training and testing sets
```{r}
#| label: regression-split

set.seed(555555)

pm25_reg_split <- initial_split(regression_data, prop = 0.75)

pm25_reg_train <- training(pm25_reg_split)
pm25_reg_test <- testing(pm25_reg_split)

# Check
nrow(pm25_reg_train)
nrow(pm25_reg_test)
```

### Recipe

```{r}
#| label: regon-recipe
pm25_reg_recipe <- recipe(mean_pm25 ~ nearest_fire_km_14d + fire_count_100km_7d + fire_count_250km_14d, data = pm25_reg_train) |>
  step_scale(all_predictors()) |>
  step_center(all_predictors())
```

### Five-fold cross-validation

```{r}
#| label: reg-cross-validation
set.seed(555555)

pm25_reg_vfold <- vfold_cv(pm25_reg_train, v = 5)
```

### KNN regression model

```{r}
#| label: knn-regression-model

knn_reg_spec <- nearest_neighbor(weight_func = "rectangular", neighbors = tune()) |>
  set_engine("kknn") |>
  set_mode("regression")
```

### Workflow

```{r}
#| label: reg-workflow
pm25_reg_workflow <- workflow() |>
  add_recipe(pm25_reg_recipe) |>
  add_model(knn_reg_spec)
```

### Tune K

First Try

```{r}
#| label: tune-knn-reg
k_reg_values <- tibble(neighbors = seq(from = 1, to = 51, by = 2))

set.seed(555555)

# tuning results
# pm25_reg_results <- pm25_reg_workflow |>
#  tune_grid(
#    resamples = pm25_reg_vfold,
#    grid = k_reg_values,
#    metrics = metric_set(rmse, mae, rsq)
#  )

# saving loaded data
# saveRDS(
#  pm25_reg_results,
#  "data/processed/pm25_knn_regression_tuning.rds"
# )

# load previously saved data
pm25_reg_results <- readRDS("data/processed/pm25_knn_regression_tuning.rds")
```

```{r}
#| label: rmse
#| output: true
regression_rmse <- pm25_reg_results |>
  collect_metrics() |>
  filter(.metric == "rmse")

head(regression_rmse)

best_reg_k <- regression_rmse |>
  arrange(mean) |>
  slice(1) |>
  pull(neighbors)

best_reg_k
```

```{r}
#| label: regression-k-visualization
#| output: true

best_reg_k_point <- regression_rmse |>
  filter(neighbors == best_reg_k)

rmse_vs_k <- regression_rmse |>
  ggplot(aes(x = neighbors, y = mean)) +
  geom_line() +
  geom_point() +
  geom_point(data = best_reg_k_point, size = 3) +
  labs(
    title = "Cross-Validation RMSE Across K Values",
    subtitle = "KNN regression using 5-fold cross-validation",
    x = "Number of Neighbours (K)",
    y = "RMSE"
  ) +
  theme_minimal()

rmse_vs_k
```

```{r}
#| label: tune-knn-reg-2
#| echo: false
k_reg_vals_refined <- tibble(neighbors = 4:10)

set.seed(555555)

# tuning results
# pm25_reg_results_refined <- pm25_reg_workflow |>
#  tune_grid(
#    resamples = pm25_reg_vfold,
#    grid = k_reg_vals_refined,
#    metrics = metric_set(rmse, mae, rsq)
#  )

# saving loaded data
# saveRDS(
#   pm25_reg_results_refined, 
#   "data/processed/pm25_knn_regression_tuning_refined.rds"
# )

# load previously saved data
pm25_reg_results_refined <- readRDS("data/processed/pm25_knn_regression_tuning_refined.rds")
```

```{r}
#| label: rmse-2
#| output: true
#| echo: false
regression_rmse_refined <- pm25_reg_results_refined |>
  collect_metrics() |>
  filter(.metric == "rmse")

best_reg_k_refined <- regression_rmse_refined |>
  arrange(mean) |>
  slice(1) |>
  pull(neighbors)

best_reg_k_refined
```

```{r}
#| label: regression-k-visualization-2
#| output: true
#| echo: false

best_reg_k_point_2 <- regression_rmse_refined |>
  filter(neighbors == best_reg_k_refined)

rmse_vs_k_refined <- regression_rmse_refined |>
  ggplot(aes(x = neighbors, y = mean)) +
  geom_line() +
  geom_point() +
  geom_point(data = best_reg_k_point_2, size = 3) +
  labs(
    title = "Zoomed in on 4 to 10",
    subtitle = "KNN regression using 5-fold cross-validation",
    x = "Number of Neighbours (K)",
    y = "RMSE"
  ) +
  theme_minimal()

rmse_vs_k_refined
```
In the broader graph, cross-validation RMSE decreased shraply as K increased from 1 reaching its lowest region around K = 4 to 10. RMSE, while slowly increasing, stabilized for larger K values.

A refined search around K = 4 to 10  region identified K = 6 as the best value for the KNN regression model. Cross-validation RMSE decreased to it minimum at K = 6 and increased toward smaller and larger values.

### Final KNN Regression

```{r}
#| label: final-knn-regression
# Final KNN model
final_knn_reg_spec <- nearest_neighbor(weight_func = "rectangular", neighbors = best_reg_k_refined) |>
  set_engine("kknn") |>
  set_mode("regression")

final_pm25_reg_workflow <- workflow() |>
  add_recipe(pm25_reg_recipe) |>
  add_model(final_knn_reg_spec)

final_knn_reg_fit <- final_pm25_reg_workflow |>
  fit(data = pm25_reg_train)
```

```{r}
#| label: test-knn-regression
#| output: true
# Test
pm25_reg_predictions <- predict(final_knn_reg_fit, pm25_reg_test) |>
  bind_cols(pm25_reg_test)

pm25_reg_predictions |> head()
```

Metrics

- RMSE: Calculates prediction error. Lower the better.
- R^2: How much of the variation the model explains. Higher the better (0-1)
- MAE: Calculates average absolute error. Lower the better.

```{r}
#| label: reg-metrics
#| output: true
# RMSE, R^2, MAE values
knn_reg_metrics <- pm25_reg_predictions |>
  metrics(truth = mean_pm25, estimate = .pred)

knn_reg_metrics
```

```{r}
#| label: knn-regression-predicted-vs-actual
#| fig-height: 5
#| fig-width: 12
#| output: true

predicted_actual_plot <- pm25_reg_predictions |>
  ggplot(aes(x = mean_pm25, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "KNN Regression: Predicted vs Actual Daily PM2.5",
    subtitle = paste("Final model with K =", best_reg_k_refined),
    x = "Actual daily mean PM2.5 (µg/m^3)",
    y = "Predicted daily mean PM2.5 (µg/m^3)"
  ) +
  theme_minimal()

predicted_actual_zoomed <- pm25_reg_predictions |>
  ggplot(aes(x = mean_pm25, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 30), ylim = c(0, 30)) +
  labs(
    title = "KNN Regression: Predicted vs Actual Daily PM2.5",
    subtitle = "Zoomed to 0–30 µg/m^3",
    x = "Actual daily mean PM2.5 (µg/m^3)",
    y = "Predicted daily mean PM2.5 (µg/m^3)"
  ) +
  theme_minimal()

predicted_actual_plot | predicted_actual_zoomed
```

## Analysis
MAE = 4.42  
This means the predictions differ from the actual daily mean PM2.5 by about 4.42 µg/m^3 on average.

RMSE = 10.5   
Being much larger than MAE suggests that there are some very large predictions errors. This can be found in the Predicted vs Actual plot where most observations are clustered near 0-30 but it doesn't predict for the extreme PM2.5 values of 100+.

R^2 = 0.156  
The model explains about 15.6% of the variation in daily mean PM2.5. So proximity and recent fire counts did contain some useful information but most PM2.5 variation remains unexplained.

Thus, the regression model captured some relationship between recent wildfire conditions and daily PM2.5 but it has limitations. It particularly tends to greatly underpredict the extreme PM2.5 events.

---

# Question 2 - Linear Regression
*Does the more flexible KNN model outperform a simple linear relationship using the same wildfire predictors?*

## Predictors
Keep the same.

---

## Linear Regression

```{r}
#| label: linear-regression
#| output: true
# without tuning
linear_reg_spec <- linear_reg() |>
  set_engine("lm")

linear_reg_recipe <- recipe(
  mean_pm25 ~
    nearest_fire_km_14d +
    fire_count_100km_7d +
    fire_count_250km_14d,
  data = pm25_reg_train
)

linear_reg_workflow <- workflow() |>
  add_recipe(linear_reg_recipe) |>
  add_model(linear_reg_spec)

linear_reg_fit <- linear_reg_workflow |>
  fit(data = pm25_reg_train)

# predict
linear_reg_predictions <- predict(linear_reg_fit, pm25_reg_test) |>
  bind_cols(pm25_reg_test)

linear_reg_metrics <- linear_reg_predictions |>
  metrics(truth = mean_pm25, estimate = .pred)

linear_reg_metrics
```

---

# KNN vs. Linear Regression

```{r}
#| label: comparison-table
#| output: true
#| echo: false
model_comparison <- tibble(
  Model = c("KNN Regression", "Linear Regression"),
  RMSE = c(
    knn_reg_metrics |> filter(.metric == "rmse") |> pull(.estimate),
    linear_reg_metrics |> filter(.metric == "rmse") |> pull(.estimate)
  ),
  MAE = c(
    knn_reg_metrics |> filter(.metric == "mae") |> pull(.estimate),
    linear_reg_metrics |> filter(.metric == "mae") |> pull(.estimate)
  ),
  R_squared = c(
    knn_reg_metrics |> filter(.metric == "rsq") |> pull(.estimate),
    linear_reg_metrics |> filter(.metric == "rsq") |> pull(.estimate)
  )
)

model_comparison |>
  knitr::kable(
    digits = 3,
    caption = "Comparison of KNN and linear regression test performance"
  )
```

The KNN and linear regression models show very similarly limited predictive performance. Linear regression had a slightly lower RMSE (10.3 µg/m^3; KNN 10.5 µg/m^3), while KNN had a lower MAE (4.42 µg/m^3; linear 4.82 µg/m^3) and a higher R^2 (0.156; linear 0.106). This indicates that KNN captured more variation and differ less from the actual value compared to linear, suggesting that wildfire and PM2.5 relationship is not strictly linear.

Nevertheless, both models were barely able to explain the overall variation. Therefore, the fire proximity and fire coutns alone are insufficient for accurate PM2.5 prediction.