---
title: "Learning Module: Auditing Code"
execute:
  warning: false
  message: false
---

This is a learning module derived from the two common mistakes I made across the files. It includes importance of going over tedious steps to verify the code works as intended and to validate the data in any datasets. (Includes AI Auditing)

```{r}
#| label: load-packages

library(tidyverse)
```

```{r}
#| label: load-data

analysis_daily <- read_csv("data/processed/analysis_daily.csv")

large_fires_clean <- read_csv("data/processed/large_fires_clean.csv", show_col_types = FALSE)
```

### Learning Objectives
By the end of this module, learners should be able to:

- distinguish bewtween code that **runs** and code that produces a **correct result**
- validate processed data against known observations
- identify inconsistencies in apparently complete data
- critically audit AI-generated or AI-assisted `

---

# Acitivty 1: No Missing Values = Correct?

Suppose the wildfire dataset contains a `Report_Date` variable.

Lets first check whether the variable contains missing values (NA).

```{r}
#| label: report-date-missing
#| output: true

large_fires_clean |>
  summarise(
    missing_report_dates = sum(is.na(Report_Date))
  )
```

There are no missing report dates.

### Question 1
Does this make it a reliable variable?

::: {.callout-tip}

### Hint
A value can exist in a data set and still be inconsistent with other information.

Try comparing the `Fire_Year` for the fire with the year contained in `Report_Date`.
:::

### Check Consistency

This code checks if any `Fire_Year` conflicts with the year in the `Report_Date`, which should be the same.

```{r}
#| label: audit-fire-date-consistency
#| output: true

large_fires_clean |>
  filter(Fire_Year != lubridate::year(Report_Date)) |>
  select(
    Fire_ID,
    Fire_Year,
    Report_Date,
    Fire_Date,
    Date_Status
  )
```

Few records contain disagreement between the fire year and report date.


::: {.callout-note}

### Guided Answer

Having zero mssing values shows that the variable is **complete** but it doesn't verify its **validity** or **consistency**.

Values should always be compared to related variables and if necessary, to external official sources.
:::

---

## AI Audit 1

An AI assistant says: `Out_Date` has no missing values, so it is safe to use directly to predictions.

What is wrong with the reasoning?


::: {.callout-note}

### Guided Answer

The AI checked **completeness** but did not check **correctness**.

A better audit:
1. checking missing values
2. comparing `Fire_Out_Year` with the year in `Out_Date`
3. investigating inconsistent records
5. documenting corrections and unresolved records
:::


---

# Activity 2: Code Runs, but is it Correct?
Suppose hourly PM2.5 measurements are being summarized into one row per monitoring station per day. (Currently, each monitoring station per day holds appx. 24 hourly PM2.5 measurements.)

We want to calculate:

- mean_pm25: average hourly PM2.5 concentration
- max_pm25: maximum hourly PM2.5 concentration

AI assistant suggests the following code:

```{r}
#| eval: false

summarise(
  mean_pm25 = mean(Concentration, na.rm = TRUE),
  max_pm25 = mean(Concentration, na.rm = TRUE)
)
```

The code executes without producing an error.

### Question 1
Is the code correct?

In other words, does the code work as intended?

::: {.callout-tip}

Hint: think about the calculation used for `mean_pm25` and `max_pm25`.

:::

### Question 2
What change should be made to the code?


::: {.callout-note}

### Guidied Answer
`max_pm25` should use `max()` rather than `mean()`

```{r}
#| eval: false
summarise(
  mean_pm25 = mean(Concentration, na.rm = TRUE),
  max_pm25 = max(Concentration, na.rm = TRUE)
)
```

The original is syntactically valid - reason it works. The mistake was in the logic. Usually, this is a common mistake that occurs when copy and pasting similar looking lines.
:::

## Validate Against a Known Observation

One useful way: compare it with a known observation from the raw dataset.

During first exploration, it was noted that the Fort St. John station recorded a high hourly PM2.5 maximum of 2112 µg/m^3 on May 19, 20223.

Given the processed dataset:

```{r}
#| label: validate-fort-st-john
#| output: true

analysis_daily |>
  filter(
    Station_Date == as.Date("2023-05-19"),
    str_detect(Station_Name, "FORT ST. JOHN")
  ) |>
  select(
    Station_Date,
    Station_Name,
    mean_pm25,
    max_pm25
  )
```

The daily mean and hourly maximum should be different. If `max_pm25` were equal to `mean_pm25`, that means the aggregation code needs to be inspected.


::: {.callout-important}
### Key Lesson
**Successful execution does not gurantee a correct analysis**

- Check whether calculated values agree with known original observations
:::

---

## AI Audit 2
Imagine a code was generated by an AI tool.

Which response is the safest?  
A. Trust the code because it ran error-free.  
B. Trust the code because the variable is named `max_pm25`  
C. Check whether the functions used actually match the intended calculations and validate the output against known observations.  
D. Remove unusually large PM2.5 values as they may be errors.   


::: {.callout-note}

### Guided Answer

**C**

Successful execution does not prove the code works as intended.

Always check the logic and compare outputs with previously known data.
:::

---

# Self-Assessment
Always verify data analysis result when there has been a change. Ask yourself:

- Did the code execute successfully?
- Does each function work as intended?
- Do the results make sense? (For example, width of a table is negative)
- Can this output be compared to a known obervation?
- Are variables consistent with other related variables?
- If AI helped generate code, did you verify?